---
title: "Error measurament simulations"
author: "Leonardo Capitani"
date: "07/17/2026"
date-modified: last-modified
execute:
  echo: true
  warning: false
  message: false
format:
      html:
          toc: true
          code-fold: true
          code-tools: true
          code-link: true
          embed-resources: true
      pdf:
             keep-tex: true
editor: visual
editor_options:
  chunk_output_type: console
---

```{r}
library(brms)
library(tidybayes)
library(tidyverse)
library(ggpubr)
```

```{r}
# Make random things reproducible
set.seed(1234)

# Bayes stuff
# Use the cmdstanr backend for Stan because it's faster and more modern than
# the default rstan. You need to install the cmdstanr package first
# (https://mc-stan.org/cmdstanr/) and then run cmdstanr::install_cmdstan() to
# install cmdstan on your computer.
options(mc.cores = 4,  # Use 4 cores
        brms.backend = "cmdstanr")
bayes_seed <- 1234

# Colors from MetBrewer
clrs <- MetBrewer::met.brewer("Java")

# Custom ggplot themes to make pretty plots
# Get Roboto Condensed at https://fonts.google.com/specimen/Roboto+Condensed
# Get Roboto Mono at https://fonts.google.com/specimen/Roboto+Mono
theme_pred <- function() {
  theme_minimal(base_family = "Roboto Condensed") +
    theme(panel.grid.minor = element_blank(),
          plot.background = element_rect(fill = "white", color = NA),
          plot.title = element_text(face = "bold"),
          strip.text = element_text(face = "bold"),
          strip.background = element_rect(fill = "grey80", color = NA),
          axis.title.x = element_text(hjust = 0),
          axis.title.y = element_text(hjust = 0),
          legend.title = element_text(face = "bold"))
}

theme_pred_dist <- function() {
  theme_pred() +
    theme(plot.title = element_markdown(family = "Roboto Condensed", face = "plain"),
          plot.subtitle = element_text(family = "Roboto Mono", size = rel(0.9), hjust = 0),
          axis.text.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank())
}

theme_pred_range <- function() {
  theme_pred() +
    theme(plot.title = element_markdown(family = "Roboto Condensed", face = "plain"),
          plot.subtitle = element_text(family = "Roboto Mono", size = rel(0.9), hjust = 0),
          panel.grid.minor.y = element_blank())
}

update_geom_defaults("text", list(family = "Roboto Condensed", lineheight = 1))
```

## Introduction

From Gelman et al. 2020 *Regression and other stories*, p. 458 and Gelman's blog [post](https://statmodeling.stat.columbia.edu/2024/04/14/simulation-to-understand-measurement-error-in-regression/) on 14.04.2024:

> Another form of incomplete data is measurement error. In the linear regression $y = a+bx+error$, measurement error in $y$ does not pose a problem, as it folds right into the error term. Suppose we would like to estimate the “underlying” model $y_i = a + bx_i + e_i$ but $y$ is itself measured with error, so that what we actually observe is $y_i^* = y_i + \eta_i$. In that case, we can combine the two models to get $y_i^* = a + bx_i + e_i + \eta_i$, and if the two error terms are independent and each with mean 0, this reduces to a linear regression of $y^*$ on $x$, which can be fit directly. **So as long as** $e_i$ **is independent of** $\eta_i$ , **you can just combine them into a single error term.**

## Simulate

First , we are going to generate fake data, following the linear relationship $y_i = a + bx_i + e_i$ with a priori fixed parameters values:

```{r}

#simulation for measurement error

set.seed(123)

n <- 1000 # number of observations

x <- runif(n, 0, 10) # predictor x with mean = 0 and sd = 10

a <- 0.2 # intercept, the mean value when x = 0
b <- 0.3 # influence of x on y
sigma <- 0.5 # also call in our model "e". It represents to what extent our model is able to predict y.

y <- rnorm(n, a + b*x, sigma)

fake <- data.frame(x,y)
```

Then fit the linear model $y_i = a + bx_i + e_i$ to the fake data:

```{r}

get_prior(formula = y ~ 1 + x, data = fake )

fit_1 <- brm(formula = y ~ 1 + x,
             family = gaussian,
             data = fake,
                prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
             iter = 2000, warmup = 1000, cores = 4, chains = 4,
             backend = 'cmdstanr',
      seed = 15)

print(fit_1)
# generate posterior predictions
pred_1 <- fit_1 |>
  add_predicted_draws(newdata = fake, ndraws = 100,
                  seed = 12345) |> mutate(model = "No Measurement error")
```

::: callout-note
**Take home message:**

The model returns exactly the fixed values we have previously determined.
:::

Then we are going to add the error measurement $\eta$ on $y$ , so $y_i^* = y_i + \eta_i$. We fixed de measurement error as $\eta = 1$. Remember that $e = 0.5$.

```{r}

sigma_y <- 1
fake$y_star <- rnorm(n, fake$y, sigma_y)

```

Then we add some error measurement on x, generated $x^*$. We fixed de measurement error as $\tau = 4$.

```{r}

sigma_x <- 4
fake$x_star <- rnorm(n, fake$x, sigma_x)
```

Then we fit again the same model, but with error measurement on $y$

```{r}

get_prior(formula = y_star ~ 1 + x, data = fake )


fit_2 <- brm(formula = y_star ~ 1 + x,
             family = gaussian,
             data = fake,
                prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
             iter = 2000, warmup = 1000, cores = 4, chains = 4,
             backend = 'cmdstanr',
      seed = 15)

print(fit_2)
# generate posterior predictions

pred_2 <- fit_2 |>
  add_predicted_draws(newdata = fake, ndraws = 100,
                  seed = 12345) |> mutate(model = "Measurement error in y")
```

Then we fit again the same model, but with error measurement on $x$

```{r}

fit_3 <- brm(formula = y ~ 1 + x_star,
             family = gaussian,
             data = fake,
                prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
             iter = 2000, warmup = 1000, cores = 4, chains = 4,
             backend = 'cmdstanr',
      seed = 15)

print(fit_3)
# generate posterior predictions

pred_3 <- fit_3 |>
  add_predicted_draws(newdata = fake, ndraws = 100,
                  seed = 12345) |> mutate(model = "Measurement error in x")
```

Then we fit again the same model, but with error measurement on $x^*$ and $y^*$.

```{r}
fit_4 <- brm(formula = y_star ~ 1 + x_star,
             family = gaussian,
             data = fake,
                prior = c(prior(normal(0, 0.2), class = Intercept),
                prior(normal(0, 0.5), class = b),
                prior(exponential(1), class = sigma)),
             iter = 2000, warmup = 1000, cores = 4, chains = 4,
             backend = 'cmdstanr',
      seed = 15)

print(fit_4)

# generate posterior predictions

pred_4 <- fit_4 |>
  add_predicted_draws(newdata = fake, ndraws = 100,
                  seed = 12345) |> mutate(model = "Measurement error in both x and y")
```

Merge the four data frames with model's predictions

```{r}
merg <-  rbind(pred_1,pred_2,pred_3,pred_4)
```

Then we plot:

```{r}

p1 <- ggplot(pred_1, aes(x = x)) +

  stat_lineribbon(aes(y = .prediction), .width = 0.95, color = "red", fill = "#D1D3D5") +
  geom_point(data = fake, aes(y = y), shape = 1, size = 1, alpha = 0.7) +
  scale_x_continuous(limits = c(-10,20), breaks = seq(from = -10, to = 20, by = 2.5))+
  scale_y_continuous(limits = c(-3, 6))+
  labs(x = "x", y = "y", color = NULL, fill = NULL,
       title = "No measurement error") +
  theme_bw(base_size = 16)


p2 <- ggplot(pred_2, aes(x = x)) +

  stat_lineribbon(aes(y = .prediction), .width = 0.95, color = "red", fill = "#D1D3D5") +
  geom_point(data = fake, aes(y = y_star), shape = 1, size = 1, alpha = 0.7) +
  labs(x = "x", y = expression(y^"*"),color = NULL, fill = NULL,
       title = "Measurement error on y") +
  scale_x_continuous(limits = c(-10,20), breaks = seq(from = -10, to = 20, by = 2.5))+
  scale_y_continuous(limits = c(-3, 6))+
  theme_bw(base_size = 16)

p3 <- ggplot(pred_3, aes(x = x_star)) +

  stat_lineribbon(aes(y = .prediction), .width = 0.95, color = "red", fill = "#D1D3D5") +
  geom_point(data = fake, aes(y = y), shape = 1,size = 1, alpha = 0.7) +
  labs(x = expression(x^"*"), y = "y",color = NULL, fill = NULL,
       title = "Measurement error on x") +
  scale_x_continuous(limits = c(-10,20), breaks = seq(from = -10, to = 20, by = 2.5))+
  scale_y_continuous(limits = c(-3, 6))+
  theme_bw(base_size = 16)

p4 <- ggplot(pred_4, aes(x = x_star)) +

  stat_lineribbon(aes(y = .prediction), , color = "red", fill= "#D1D3D5", .width = 0.95) +
  geom_point(data = fake, aes(y = y_star),shape = 1,  size = 1, alpha = 0.7) +
  labs(x = expression(x^"*"), y =expression(y^"*"),color = NULL, fill = NULL,
       title = "Measurement error in both x and y") +
  scale_x_continuous(limits = c(-10,20), breaks = seq(from = -10, to = 20, by = 2.5))+
  scale_y_continuous(limits = c(-3, 6))+
  theme_bw(base_size = 16)


ggarrange(p1, p2, p3, p4,

          ncol = 2, nrow = 2)


```

We can compare the regression coefficients:

```{r}

tidy_inter_1 <-  bayesplot::mcmc_intervals_data(fit_1,
              prob = 0.66, # 66% intervals
              prob_outer = 0.95,#95% intervals
           pars = vars(-contains(c("prior_", "lprior", "lp__")))) |> mutate(model = "No Measurement error")

tidy_inter_2 <-  bayesplot::mcmc_intervals_data(fit_2,
              prob = 0.66, # 66% intervals
              prob_outer = 0.95,#95% intervals
           pars = vars(-contains(c("prior_", "lprior", "lp__")))) |> mutate(model = "Measurement error in y")

tidy_inter_3 <-  bayesplot::mcmc_intervals_data(fit_3,
              prob = 0.66, # 66% intervals
              prob_outer = 0.95,#95% intervals
           pars = vars(-contains(c("prior_", "lprior", "lp__")))) |> mutate(model = "Measurement error in x")

tidy_inter_4 <-  bayesplot::mcmc_intervals_data(fit_4,
              prob = 0.66, # 66% intervals
              prob_outer = 0.95,#95% intervals
           pars = vars(-contains(c("prior_", "lprior", "lp__")))) |> mutate(model = "Measurement error in both x and y")


me <- rbind(tidy_inter_1, tidy_inter_2 , tidy_inter_3, tidy_inter_4) |> filter(parameter != "Intercept")

levels(me$parameter)[levels(me$parameter) == "b_x_star"] <- "b_x" # rename the b parameter where we use x_star

pos <- position_nudge(y = case_when(me$model == "No Measurement error" ~ 0,
                                 me$model  == "Measurement error in y" ~ 0.05,
                                  me$model  == "Measurement error in x" ~ 0.1,
                                    me$model  == "Measurement error in both x and y" ~ 0.15))


ggplot(me, aes(x = m, y = factor(parameter), color = model)) +
   geom_pointinterval(aes(xmin = l, xmax = h), position = pos, size = 8)+
   geom_pointinterval(aes(xmin = ll, xmax = hh), position = pos, size = 2)+
   geom_point( position = pos, color = "black")+
  theme_bw(base_size = 18) +
  scale_color_brewer(palette = "Set1")+
  labs(x = NULL, y = NULL)+
  scale_x_continuous(breaks = seq(from = 0, to = 2, by = 0.15))

```

::: callout-important
## Take home messages:

**The slope** $b$ **changes with measurement error in** $x$ **but not in** $y$**.**

When you have measurement error in $x$, two things happen to attenuate $b$—that is, to pull the regression coefficient toward zero.

1.  First, if you spreading out $x$ but keep $y$ unchanged, this will reduce the slope $b$ of $y$ on $x$.
2.  Second, when you add noise to $x$ you’re changing the ordering of the data, which will reduce the strength of the relationship.
:::

## Fitting a Bayesian regression with error measuraments on predictors $x$

Please see [here](https://mbjoseph.github.io/posts/2018-12-25-errors-in-variables-models-in-stan/).
