---
title: "Empirical Dynamic Modelling: an introduction"
author: "Leonardo Capitani"
date: "07/22/2026"
date-modified: last-modified
execute:
  echo: true
  warning: false
  message: false
format:
      html:
          toc: true
          code-fold: true
          code-tools: true
          code-link: true
          embed-resources: true
      pdf:
             keep-tex: true
editor: visual
editor_options:
  chunk_output_type: console
bibliography: references.bib
csl: "ecology-letters.csl"
---

```{r load_libraries}
library(rEDM)
library(tidyverse)
```

## Why not just fit a regression?

You've got a population time series and you want to know what drives it. The reflex move is a regression, or maybe an ARIMA model. Here's the problem with that:

-   Regression assumes a **fixed** relationship: a 1-unit change in $X$ always buys you the same change in $Y$.
-   Ecological systems are usually **state-dependent**: the effect of a predator on its prey can flip sign depending on whether prey are scarce or abundant, whether the system is near a tipping point, and so on.
-   This gives you "mirage correlations" — correlations between two variables that appear, vanish, and reverse over time, even though the two variables are genuinely causally linked the whole time [@chang_empirical_2017].

**Q: So what do I do instead?**
A: You stop trying to write down the equations, and instead let the data reconstruct the dynamics for you. That's Empirical Dynamic Modelling (EDM) — a non-parametric, nonlinear alternative to regression, built for exactly this kind of system [@chang_empirical_2017].

## The core idea: rebuilding an attractor from one time series

Here's the counterintuitive part.

-   Say you only measured *one* species in a multi-species food web. Surely you've lost all the information about the other species?
-   Not quite: if that species interacts with the rest of the system, its own trajectory carries a "shadow" of everything it interacts with.
-   **Takens' theorem** says you can recover that shadow using nothing but lagged copies of your one time series.

**Q: How, mechanically?**
A: You build an embedding — a vector of $E$ lagged values of your series:

$$
\vec{x}_t = (x_t,\ x_{t-\tau},\ x_{t-2\tau},\ \dots,\ x_{t-(E-1)\tau})
$$

-   $E$ = the **embedding dimension** — how many lags you use.
-   $\tau$ = the **time delay** between lags (I'll use $\tau=1$ throughout, one time step).
-   Plot these vectors in $E$-dimensional space and you get a "shadow manifold" — a reconstruction of the system's attractor that shares its essential shape (topology) with the real thing.

You never see the real attractor. You don't need to: the shadow is enough to forecast from.

## Step 1 — Simplex projection: forecasting by analogy

The simplest EDM forecasting method is **Simplex projection**, and the logic is just "find what happened last time things looked like this":

-   Embed the series into $E$ dimensions.
-   For the point you want to forecast, find its $E+1$ nearest neighbours in the embedded space (using only points from the past, i.e. your "library").
-   Forecast = a distance-weighted average of what those neighbours did next.

That's it — no equations, no parameters to fit, just pattern-matching in the reconstructed state space.

**Q: What counts as a good population model to try this on?**
A: A chaotic one. Robert May showed back in 1976 that ridiculously simple population models (like a discrete logistic map) can produce dynamics so complex they look like noise — but they aren't. I'll use `TentMap`, a textbook chaotic map bundled with `rEDM`, as a stand-in for a "simple but chaotic" single-species population.

### Picking $E$

I don't get to pick $E$ by intuition — I let forecast skill tell me. I split the series into a **library** (used to search for neighbours) and a **prediction** set (held out, forecasted), then try a range of $E$ values:

```{r embed_dimension}
lib_range  <- "1 100"
pred_range <- "201 500"

E_test <- EmbedDimension(
  dataFrame = TentMap, lib = lib_range, pred = pred_range,
  columns = "TentMap", target = "TentMap", showPlot = FALSE
)

E_test |>
  ggplot(aes(x = E, y = rho)) +
  geom_line(color = "#1B7837") +
  geom_point(size = 2, color = "#1B7837") +
  labs(x = "Embedding dimension (E)", y = expression(rho~"(forecast skill)"),
       title = "Simplex forecast skill across embedding dimensions") +
  theme_bw(base_size = 14)

best_E <- E_test$E[which.max(E_test$rho)]
best_E
```

**Q: What am I looking for in that plot?**
A: A peak. Too few dimensions and the reconstruction folds different states on top of each other (bad neighbours); too many and I'm fitting noise. The peak — here $E = `r best_E`$ — is my best embedding.

## Is it chaos, or is it just noise?

This question is where EDM started: Sugihara and May showed that nonlinear forecasting like this could tell real deterministic chaos apart from measurement error in ecological and epidemiological time series (think measles, chickenpox) — cases where a simple linear analysis just sees "noise" [@sugihara_may_1990]. Here's the diagnostic: forecast the series at increasing horizons ($T_p = 1, 2, 3, \dots$) and watch what happens to forecast skill.

-   **Chaotic** systems: skill decays *smoothly* with horizon (small errors compound — the "butterfly effect").
-   **Noise**: skill is already bad at $T_p=1$ and stays flat and low.

```{r chaos_vs_noise}
horizon_chaos <- PredictInterval(
  dataFrame = TentMap, lib = lib_range, pred = pred_range,
  E = best_E, columns = "TentMap", target = "TentMap", showPlot = FALSE
) |> mutate(series = "Chaotic (TentMap)")

horizon_noise <- PredictInterval(
  dataFrame = TentMapNoise, lib = lib_range, pred = pred_range,
  E = best_E, columns = "TentMap", target = "TentMap", showPlot = FALSE
) |> mutate(series = "Noise (TentMapNoise)")

bind_rows(horizon_chaos, horizon_noise) |>
  ggplot(aes(x = Tp, y = rho, color = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("#1B7837", "grey60")) +
  labs(x = "Prediction horizon (Tp, steps ahead)", y = expression(rho),
       color = NULL, title = "Forecast skill vs. prediction horizon") +
  theme_bw(base_size = 14)
```

**Q: Why does this matter for a real dataset?**
A: If your population's forecast skill degrades gradually like the chaotic curve, you have a *deterministic but unpredictable-in-the-long-run* system — a completely different management problem than pure noise (where more monitoring effort won't buy you better forecasts).

## Step 2 — S-map: does the "rulebook" change with the state of the system?

Simplex gives you a forecast; **S-map** [@sugihara_1994] (sequential locally weighted global linear map) gives you something more diagnostic: it fits a *local* linear model around each point, weighting neighbours by

$$
w(d) = \exp\!\left(-\theta \frac{d}{\bar d}\right)
$$

-   $d$ = distance to a neighbour, $\bar d$ = average distance.
-   $\theta = 0$: every point gets equal weight → one **global** linear model (this is just standard linear autoregression).
-   $\theta > 0$: nearby points matter more → the "rules" are allowed to change across the state space (**state-dependence**, i.e. nonlinearity).

**Q: How does this tell me the system is nonlinear?**
A: If forecast skill *improves* as $\theta$ increases above 0, a single fixed rule can't explain your data — the best-fitting local model changes depending on where you are on the attractor. That is itself evidence of nonlinear, state-dependent dynamics.

```{r smap_theta}
theta_test <- PredictNonlinear(
  dataFrame = TentMap, lib = lib_range, pred = pred_range,
  E = best_E, columns = "TentMap", target = "TentMap", showPlot = FALSE
)

theta_test |>
  ggplot(aes(x = Theta, y = rho)) +
  geom_line(color = "#1B7837") +
  geom_point(size = 2, color = "#1B7837") +
  labs(x = expression(theta~"(local weighting / nonlinearity)"), y = expression(rho),
       title = "S-map: does nonlinearity improve forecast skill?") +
  theme_bw(base_size = 14)
```

The peak sits well above $\theta = 0$: a single global linear rule underfits this system — exactly what I'd expect from a chaotic map, and exactly the pattern you should look for before assuming a linear model is good enough for your own data.

## Trying it on a real population

`rEDM` ships a real dataset I like a lot for this: monthly counts of *Thrips imaginis* (an insect) in South Australia, alongside rainfall and temperature — a genuine, noisy, real-world ecological time series, not a toy.

```{r thrips_setup}
data(Thrips)
Thrips |> select(Year, Month, Thrips_imaginis) |> head()
```

**Q: Same two questions as before — what E, and is it nonlinear?**

```{r thrips_edm}
lib_thrips <- paste("1", nrow(Thrips))

E_thrips <- EmbedDimension(
  dataFrame = Thrips, lib = lib_thrips, pred = lib_thrips,
  columns = "Thrips_imaginis", target = "Thrips_imaginis", showPlot = FALSE
)
best_E_thrips <- E_thrips$E[which.max(E_thrips$rho)]

theta_thrips <- PredictNonlinear(
  dataFrame = Thrips, lib = lib_thrips, pred = lib_thrips,
  E = best_E_thrips, columns = "Thrips_imaginis", target = "Thrips_imaginis",
  showPlot = FALSE
)

theta_thrips |>
  ggplot(aes(x = Theta, y = rho)) +
  geom_line(color = "#1B7837") +
  geom_point(size = 2, color = "#1B7837") +
  labs(x = expression(theta), y = expression(rho),
       title = paste0("Thrips imaginis, E = ", best_E_thrips)) +
  theme_bw(base_size = 14)
```

Forecast skill is far from perfect (real insect counts are noisy), but it clearly rises above the linear ($\theta=0$) baseline — a real population, not a simulation, showing the same nonlinear signature I built into the chaotic toy example above.

## Step 3 — S-map coefficients: reading off interaction strength

So far I've only used S-map's *forecast skill*. But since S-map fits an actual local linear model at every point, I can also look at its **coefficients** — and if I embed more than one species together, those coefficients are interpretable as time-varying interaction strengths (essentially, a Jacobian): "how much does a change in species $j$ move species $i$, right now, given the current state of the system?" [@sugihara_1994]

I'll use `block_3sp`, a bundled 3-species dataset already lagged for me:

```{r block3sp_setup}
data(block_3sp)
species_cols <- c("x_t", "y_t", "z_t")
block_3sp |> select(time, all_of(species_cols)) |> head()
```

```{r smap_coefficients}
smap_out <- SMap(
  dataFrame = block_3sp, lib = "1 100", pred = "101 190", E = 3,
  columns = species_cols, target = "x_t", embedded = TRUE, theta = 2,
  showPlot = FALSE
)

smap_out$coefficients |>
  rename(effect_of_x = 3, effect_of_y = 4, effect_of_z = 5) |>
  select(time, effect_of_x, effect_of_y, effect_of_z) |>
  pivot_longer(-time, names_to = "predictor", values_to = "coefficient") |>
  ggplot(aes(x = time, y = coefficient, color = predictor)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("#1B7837", "#762A83", "#2166AC"),
                      labels = c("effect of x on x", "effect of y on x", "effect of z on x")) +
  labs(x = "Time", y = "S-map coefficient", color = NULL,
       title = "Time-varying effect of x, y and z on species x") +
  theme_bw(base_size = 14) +
  theme(legend.position = "bottom")
```

**Q: why not just fit one regression coefficient per pair of species and be done with it?**
A: Because in a nonlinear system that single number is a lie by omission — the true effect of $y$ on $x$ isn't constant, it depends on the state you're in. The plot above *is* that dependence, made visible: the "effect of $y$ on $x$" line isn't flat, it swings around zero as the system moves through different parts of its attractor.

## Step 4 — Multiview embedding: letting the data pick the best variables

Everything above used one fixed embedding. **Multiview** asks: what if I don't have to commit to one? Ye and Sugihara's approach is refreshingly simple [@ye_2016]:

-   Build *every* candidate embedding of dimension $E$ you can make out of your observed variables (including different lags).
-   Rank them by in-sample forecast skill.
-   Forecast using the top-ranked ones, and average the results.

**Q: doesn't that risk overfitting — cherry-picking whatever worked in-sample?**
A: A little, which is why you average several good embeddings rather than keeping only the single best one — that's what tames the cherry-picking. Let's check whether it actually beats a plain univariate forecast on the same 3-species system:

```{r multiview}
uni_out <- Simplex(
  dataFrame = block_3sp, lib = "1 100", pred = "101 195",
  E = 3, columns = "x_t", target = "x_t", showPlot = FALSE
)
uni_rho <- ComputeError(uni_out$Observations, uni_out$Predictions)$rho

mv_out <- Multiview(
  dataFrame = block_3sp, lib = "1 100", pred = "101 195",
  E = 3, columns = species_cols, target = "x_t", showPlot = FALSE
)
mv_rho <- ComputeError(mv_out$Predictions$Observations, mv_out$Predictions$Predictions)$rho

tibble(
  method = c("Univariate Simplex (x only)", "Multiview (x, y, z)"),
  rho = c(uni_rho, mv_rho)
)
```

Multiview comes out ahead, using nothing I hadn't already measured — just letting the data decide which lagged combinations of $x$, $y$ and $z$ are worth listening to, instead of me guessing.

## Take-home

-   Linear correlation-based methods assume a fixed relationship. Ecological systems often don't have one.
-   EDM sidesteps that by reconstructing the system's attractor directly from lagged values of a time series (Takens' theorem) — no equations required.
-   **Simplex projection** forecasts by analogy to the nearest neighbours in that reconstruction; sweeping $E$ tells you the reconstruction's dimensionality, and sweeping $T_p$ tells you chaos from noise [@sugihara_may_1990].
-   **S-map** [@sugihara_1994] checks whether a *local, state-dependent* model beats a single global linear one — the diagnostic for nonlinearity — and its coefficients double as time-varying interaction strengths once you embed more than one variable.
-   **Multiview** [@ye_2016] lets forecast skill, not intuition, pick which combination of variables and lags to trust.
-   Everything here used one variable at a time (or, for S-map coefficients and Multiview, a handful measured together). The natural next question is *interaction between* variables — does species $A$ actually drive species $B$? That's what convergent cross mapping is for, next notebook.

## Further reading

-   @chang_empirical_2017 — a hands-on `rEDM` walkthrough I leaned on heavily here.
-   [Time Series Analysis Handbook, ch. 5 — Simplex and S-map projections](https://phdinds-aim.github.io/time_series_handbook/05_SimplexandSmapProjections/05_Empirical%20Dynamic%20Modelling%20%28Simplex%20and%20SMap_Projections%29.html) — the Python-flavoured version of everything above, with more detail on multivariate embeddings.
-   [The official `rEDM` package vignette](https://ha0ye.github.io/rEDM/articles/rEDM.html) — covers the same ground as this notebook (Simplex, S-map, S-map coefficients, Multiview, CCM) using the package's earlier API; a good second reference if you get stuck on syntax.
-   @sugihara_may_1990 — the founding EDM paper: using nonlinear forecasting to tell chaos from measurement error in real epidemiological time series.
-   @sugihara_1994 — introduces S-map.
-   @dixon_1999 — an early, high-stakes real-world validation: forecasting larval reef-fish supply, not just toy chaotic maps.
-   @sugihara_2012 — introduces Convergent Cross Mapping; the full story is in the [next notebook](../edm/CCM_example.qmd).
-   @ye_2016 — introduces Multiview embedding.
-   @ye_distinguishing_2015 — the time-lag extension to CCM, also covered in the [next notebook](../edm/CCM_example.qmd).
